GO-terms enriched in heatmap gene clusters for mouse and \textit{Eimeria falciformis}

### 2016-06-03 additional_files/tex/SI_clusters/clusterTablesCombined.tex
Mainly gne_list changes --> orthologs from ToxoDB, and some text in manuscipt draft (including some obvious errors in results part)

%\usepackage[version=3]{mhchem} % Package for chemical equation typesetting
%\usepackage{siunitx} % Provides the \SI{}{} and \si{} command for typesetting SI units
\usepackage{graphicx} % Required for the inclusion of images
\usepackage[section]{placeins}
\usepackage[hscale=0.8,vscale=0.8]{geometry}
%\usepackage{natbib} % Required to change bibliography style to APA
%\usepackage{amsmath} % Required for some math elements

@@

\section{Results for \textit{E. falciformis}}
  Cluster 1, BP, Ef.
  Overrepresented GO biological process terms among parasite genes which are 
  highly abundant on day 7 p.i.. Compared to cluster 2, which also have a high
  abundance on day 7, these genes cluster more closely with genes in cluster 4, 
  which are highly expressed in oocysts.\\
  %\newpage
  \input{../cluster1_BP_Ef_Day7Up_1.tex}

  Cluster 2, BP, Ef.
  Overrepresented GO biological process terms among parasite genes which are 
  highly abundant on day 7 p.i.. Compared to cluster 1, which is also upregulated 
  on day 7 p.i., these genes cluster more closely with cluster 6, which contains
  genes highly abundant in sporozoites.\\
  \input{../cluster2_BP_Ef_Day7Up_2.tex}

  Cluster 1, MF, Ef.\\
  Overrepresented GO molecular function terms among parasite genes which are 
  highly abundant on day 7 p.i.. Compared to cluster 2, which also have a high
  abundance on day 7, these genes cluster more closely with genes in cluster 4, 
  which are highly expressed in oocysts.\\
  \input{../cluster1_MF_Ef_Day7Up_1.tex}

  Cluster 2, MF, Ef.
  Overrepresented GO molecular function terms among parasite genes which are 
  highly abundant on day 7 p.i.. Compared to cluster 1, which is also upregulated 
  on day 7 p.i., these genes cluster more closely with cluster 6, which contains
  genes highly abundant in sporozoites.\\
  \input{../cluster2_MF_Ef_Day7Up_2.tex}

  Cluster 3, BP, Ef.
  Overrepresented GO biological process terms among parasite genes which have a low
  abundance in sporulated oocysts and sporozoites. In mouse stages, i.e., day 3, 5, and 7 samples
  there is a tendency to upregulation among these genes.\\
  \input{../cluster3_BP_Ef_OocSporDown.tex}

@@

  Cluster 4, BP, Ef.
  Overrepresented GO biological process terms among parasite genes which are highly abundant in
  sporulated oocysts. These genes cluster most closely with cluster 1, in which genes are highly
  abundant on day 7 p.i.. In cluster 4, genes have below average abundance in all day 3 and 5 samples.\\
  \input{../cluster4_BP_Ef_OocystsUp.tex}

@@

  Cluster 5, BP, Ef.
  Overrepresented GO biological process terms among parasite genes which have low abundance
  in sporulated oocysts, sporozoites and on day 7 p.i.. These genes seem to increase in abundance
  upon infection and decrease late in infection, i.e., on day 7.\\
  \input{../cluster5_BP_Ef_OoSpoDown_day5Up.tex}

@@

  Cluster 6, BP, Ef.
  Overrepresented GO biological process terms among parasite genes which are highly abundant in sporozoites
  but mainly downregulated in sporulated oocysts, apart from a few genes in the cluster.
  This cluster seems to distinguish sporozoites from sporulated oocysts and have an average mRNA abundance in
  all other samples.\\
  \input{../cluster6_BP_Ef_SporozoitesUp.tex}

@@

  Cluster 7, BP, Ef.
  Overrepresented GO biological process terms among parasite genes which have low abundance in sporulated oocysts.
  In sporozoites and early in infection, i.e., day 3 and 5 p.i., some of these genes are highly abundant. On day
  7 p.i. these mRNAs have a below average abundance.\\
  \input{../cluster7_BP_Ef_Day7OocyDown_SporoUP.tex}

@@

\section{Results for mouse}
  Cluster 1, BP, Mm
  Overrepresented GO biological process terms among mouse genes which have low abundance pre infection and 
  become more abundant in most day 3 and 5 samples from immunecompetent mice, but less in immune deficient Rag1-/- mice. 
  The upregulation trend is most clear in three first infection day 5 samples, which also form a separate
  sample cluster. On day 7, most of these genes are not differentially abundant, or below average, i.e., they
  seem to peak on day 5 p.i..\\
  \input{../cluster1_BP_Mm_EarlyUp.tex}

@@

  Cluster 2, BP, Mm.
  Overrepresented GO biological process terms among mouse genes which are downregulated on day 7 p.i. and
  upregulated pre infection, i.e., these genes become more and more downregulated during infection.\\
  \input{../cluster2_BP_Mm_7dpiUp.tex}

@@

  Cluster 3, BP, Mm.
  Overrepresented GO biological process terms among mouse genes which are upregulated pre infection and
  downregulated on day 7 p.i., except in one NMRI second infection sample, possibly indicating no or
  poor infection in this sample. As in cluster 2, these mRNAs are most highly abundant before infection and
  become less abundant over the course of infection.\\
  \input{../cluster3_BP_Mm_7dpiDown.tex}

@@

  Cluster 4, BP, Mm.
  Overrepresented GO biological process terms among mouse genes which are highly abundant on day 7 p.i.. 
  There is overall no clear difference between pre infection and days 3 or 5 among these genesi. A slight
  tendency to higher abundance in second infection samples can be seen among these genes.\\
  \input{../cluster4_BP_Mm_allInfDown.tex}

>>>>

%\usepackage[version=3]{mhchem} % Package for chemical equation typesetting
%\usepackage{siunitx} % Provides the \SI{}{} and \si{} command for typesetting SI units
\usepackage{graphicx} % Required for the inclusion of images
\usepackage[section]{placeins}
\usepackage[hscale=0.8,vscale=0.85]{geometry}
%\usepackage{natbib} % Required to change bibliography style to APA
%\usepackage{amsmath} % Required for some math elements

@@

\section{Results for \textit{E. falciformis}}
Cluster 1, BP, Ef.
Overrepresented GO 'biological process' terms among parasite mRNAs which are 
highly abundant on day 7 p.i. only. Cluster 1 groups closely with cluster 4,
in which mRNAs are highly abundant in oocysts. The cluster contains 201 genes of which 
96 have GO annotations.\\

\newline
The eight first GO terms in this list are supported by the same three genes. 
Unintuitively for a protozoan organism, seven out of eight of these terms are associated with 
wound healing and blood coagulation.
However, orthologs in ToxoDB pointed towards \textit{Eimeria} spp. and \textit{Toxoplasma gondii} 
proteins annotated as micronemal protein 4, MIC4, related proteins and more generally, PAN 
domain containing proteins. The PAN domain is found in plasminogen
/hepatocyte growth factor family and in coagulation factor XI family, explaining the GO terms. 
Of interest here, the same domain has also been reported to be common in 
nematodes such as \textit{Caenorhabditis elegans}, however the
function is not understood. (Thordai99)
In \textit{Eimeria} spp. and the apicomplexan parasite \textit{Hammondia hammondi}, orthologs
are annotated as micronemal protein MIC4, a microneme protein reported by, e.g., the Tomley group
(Brown01).
Later publications on \textit{T. gondii} (Marchant12) also associate these domains and proteins with 
micronemes and therefore invasion. In our case, this is a peculiar result, since the enrichment appears on
day 7 p.i.. A possible role at this timepoint is suggested by work on the fungi 
\textit{Sclerotinia sclerotiorum} where Yu et al. reported an important role for PAN domain
proteins in cell wall integrity (Yu12). This role for MIC proteins has to our knowledge 
not been investigated in apicomplexan parasites. 
The other two GO terms in the cluster of day seven upregulated genes are DNA replication and DNA 
replication initiation, which most likely reflects late stage schizogony or gamete formation.
%\newpage
\input{../cluster1_BP_Ef_Day7Up_1.tex}

Cluster 1, MF, Ef.\\
Overrepresented GO molecular function terms among parasite genes which are 
highly abundant on day 7 p.i.. Compared to cluster 2, which also have a high
abundance on day 7, these genes cluster more closely with genes in cluster 4, 
which are highly expressed in oocysts.\\
%%%
Four GO molecular functions are enriched in cluster 1 (adj. p-value = 0.14), 
supported by ten different genes, i.e., there is overlap in the genes. 
Carbon metabolism genes are represented by 6-phosphogluconate dehydrogenase and glycogen phosphorylase 
family protein 1. UDP-glucose 4-epimerase and amiloride-sensitive amine oxidase are 
reported as upregulated in gametocytes in \textit{E. tenella} by RNA-seqi (Walker15) and 
suggested by those authors to play a role in cyst wall synthesis.
\input{../cluster1_MF_Ef_Day7Up_1.tex}

Cluster 2, BP, Ef.
Overrepresented GO biological process terms among parasite genes which are 
highly abundant on day 7 p.i.. Compared to cluster 1, this cluster contains genes 
which are highly abundant both on day 7 p.i. and in sporozoites (cluster 6).
In total the cluster contains 271 genes of which 107 have GO annotations.\\
\newline
The term 'single-organism process' is enriched and is the GO-term containing the largest
number of genes in cluster 2 (38). The two 'single-organism' terms have a complete overlap of genes.
All genes contributing to other GO terms in this list, except those terms indicating motility 
(top 3 GO terms), are also contained in the 38 'single-organism' genes.
Therefore, although annotations such as purine/nucleoside/nucleotide metabolic process 
dominate, the genes responsible for those enrichments are all contained in the two 'single organism'
terms.
A closer look at these 38 genes reveals 12 genes annotated as either dynein, kinesin or tubulin chains or 
proteins, indicating activity related to parasite motility. (((((((((( COMMENT ))))))))))
Also potentially linked to motility is the 
occurence of growth arrest specific protein 8, Gas8, which in the mouse has been reported to be highly 
expressed in the testes and important for mouse sperm function (Yeh02). Other genes among the 38 indicate
carbon fixation (glycolysis/gluconeogenesis) or different conversions of nucleoside phosphates. 
In addition, a Ras family protein, RNA polymerase II transcription initiation factor and 
Sec23 and Sec24 were among orthologs identified in \textit{E. falciformis} cluster 2.
%%%
In the list of enriched GO 'molecular functions', 'microtubule motor activity' also indicate 
parasite movement. Under the term 'protein binding', genes annotated as dynein and kinesin 
are different from the genes annotated as dynein and kinesin in the list for 
GO 'biological process'. 
Orthologs to genes responsible for other GO term enrichments in cluster 2 also indicate motility
with annotations including 'flagellar' or 'intraflagellar'.
In addition, there are two 'EF-hand 
domain containing proteins' annotations as well as direct naming of such proteins: caltractin and 
centrin-1, and troponin. Caltractin and centrin-1 are associated with the centrosome 
and structure and function of microtubuli in mammals, and troponin is linked to muscle function 
(UniProt). 
\input{../cluster2_BP_Ef_Day7Up_2.tex}

Cluster 2, MF, Ef.
Overrepresented GO molecular function terms among parasite genes which are 
highly abundant on day 7 p.i.. Compared to cluster 1, which is also upregulated 
on day 7 p.i., these genes cluster more closely with cluster 6, which contains
genes highly abundant in sporozoites.\\
%%%

\input{../cluster2_MF_Ef_Day7Up_2.tex}

Cluster 3, BP, Ef.
Overrepresented GO biological process terms among parasite genes which have a low
abundance in sporulated oocysts and sporozoites. In mouse stages, i.e., day 3, 5, and 7 samples
there is a tendency to upregulation among these genes. In total the cluster contains 204 genes of which 
121 have GO annotations, and 83 do not.\\
 \\
\input{../cluster3_BP_Ef_OocSporDown.tex}

@@

Cluster 4, BP, Ef.
Overrepresented GO biological process terms among parasite genes which are highly abundant in
sporulated oocysts. These genes cluster most closely with cluster 1, in which genes are highly
abundant on day 7 p.i.. In cluster 4, genes have below average abundance in all day 3 and 5 samples.
In total the cluster contains 316 genes of which 157 have GO annotations, and 159 do not.\\
 \\

  Serine hydroxymethyltransferase --> (L-)serine metabolism...(?)
  Cystathionine beta-synthase -->  converts (L-)homocysteine to cystathionine using L-serine
  Acyl-CoA dehydrogenase    --> beta-oxidation in mitochondria
  MmgE/PrpD family domain -->  PrpD important for propionate catabolism in E. coli (see Sushi-domain)
  Thrombospondin type 1 domain-containing... --> microneme protein implicated in adhesion
                                                MIC proteins in Tomley et al (2001)
  Sushi domain (scr repeat) domain

  Sushi domain protein: complement binding in mammals
In Pf microneme localised (OKeeffe05) and suggested GPI anchored. Expressed in merozoites (erythrocyte invasion
stage) --> in Eimeria different, since oocyst high expression?
MmgE/PrpD is important for propionate catabolism in the 2-methylcitric acid cycle mainly in E. coli and
other bacteria (UniProt). Propionyl-CoA can be produced from propanoate and is reported to
be toxic in... not sure about the statements in the databases here. Check these sites:
http://www.ebi.ac.uk/intenz/query?cmd=SearchEC&ec=2.8.3.1
https://en.wikipedia.org/wiki/Propionate_CoA-transferase
http://www.llamp.net/?q=2-Methylcitrate%20cycle
...and these publications:
http://www.ncbi.nlm.nih.gov/pubmed/16151139
http://www.ncbi.nlm.nih.gov/pubmed/10482501
http://www.ncbi.nlm.nih.gov/pubmed/19767840

Propionate is one of two most abundant small-chain fatty acids 
in the gut along with butyrate. Both fatty acids are largely produced as degradation products from food
by commensal bacteria (Sun13).

## OLD CLUSTERING REFS???
\input{../cluster4_BP_Ef_OocystsUp.tex}

@@

ATPase or histidine kinase or DNA gyrase B or HSP90:
Gyrase has also been found in Plasmodium (Wiki) but is commonly known from bacteria, 
where it contributes to negative supercoiling of DNA. Many antibiotics act on gyrases
according to Wiki.

Patched family protein - patched is a membrane receptor which transduces hedgehog signals 
in e.g. humans (Pfam). Conserved pathway, involved in embryonic development and 
tissue maintenance. --> Not really able to find references for protozoa. Plasmodium
paper (doi:  10.1186/s12936-015-0562-1) states possible crosstalk between parasite
protein and host hedgehog...

LCCL (Limulus clotting factor C, Coch-5b2 (Cochlin) and Lgl1) -->
In metazoa, this clotting factor has been associated with other protein domains,
including discoidin lectin domains (Pfam entry for LCCL domain).
  F5/8 type C domain - main annotation is coagulation factor. Perhaps more 
relevant here, a similar domain has been found in slime molds (Dictyostelium discoideum)
where it is part of the adhesion protein discoidin (Pfam entry for "Discoidin domain")

Alkyl sulfatase --> cleaves organic sulfure: "Usually inorganic sulfate is 
available and is transformed through a cascade of reactions to yield the 
high-energy intermediate 3′-phosphoadenosine-5′-phosphosulfate (PAPS). 
Ultimately, sulfur is incorporated into the essential amino acids 
cysteine and methionine. In case inorganic sulfate is unavailable, organisms 
are forced to express other sulfur metabolizing enzymes, such as alkyl sulfatases,
which are able to hydrolytically cleave inorganic sulfate from organic sulfate 
esters" (doi:  10.1007/s00253-013-5438-0)

  Aminopeptidase N --> Peptide cleaving, in humans mainly after digestion by gastric
and pancreatic proteases. Curious: "Found to cleave antigen peptides bound to major 
histocompatibility complex class II molecules of presenting cells" (UniProt for
Homo sapiens "amonipeptidase N").
  X-Pro dipeptidyl-peptidase domain --> aminopeptidase activity, serine targeting
(InterPro, IPR008252). The Toxo annotation for CocE/NonD are serine esterases specifically
enabling using cocaine as carbon source (InterPro).

  Aminotransferase --> most likely linked to amino acid biosynthesis (UniProt for 
Drosophila)
  Nicalin-1 --> binds Nomo. Important for embryonic development in vertebrates (PMID:
15257293). Stated on Zebrafish DB that it's important for egg hatching, but it
has the same reference as above. 
  Toxo: M28 family peptidase --> contains aminopeptidases (Merops Sanger entry
  Aspartyl protease --> specific protease, cleaves in one step, most commonly
found in stomach of mammals (PMID:1455179, review on...).
  Glucose-methanol-choline, GMC, oxidoreductase --> family of flavoenzymes, which
contains, e.g., glucose dehydrogenases (DOI: 10.1016/0022-2836(92)90992-S)
  Mannitol-1-phosphate dehydrogenase AKA fructose 6-phosphate reductase -->

  Glucose-6-phosphate 1-dehydrogenase --> oxidizes glucose-6-phosphate and reduces
NADP+ to NADPH. 
  D-3-phosphoglycerate dehydrogenase --> PGDH catalyses an early step in the biosynthesis of L-serine by converting 
D-3-phosphoglyceric acid to hydroxypyruvic acid phosphate (HPAP), utilising 
NAD+ as a coenzyme [PMID: 19924905, PMID: 22023909].(InterPro)
     transhydrogenase --> oxidoreductase family, NADH production
OR   alanine dehydrogenase --> L-alanine + H2O + NAD+ <-> pyruvate + NH3 + NADH + H+
    Protein NNT (nicotinamide nucleotide transhydrogenase) -->
Identified in Plasmodium where it seems to be expressed in gametocytes.
(Thesis: http://hdl.handle.net/1860/477). For human: "The enzyme couples hydride 
transfer between NAD(H) and NADP(+) to proton translocation across the inner 
mitochondrial membrane. Under most physiological conditions, the enzyme uses energy 
from the mitochondrial proton gradient to produce high concentrations of NADPH. 
The resulting NADPH is used for biosynthesis and in free radical detoxification."
(NCBI, Gene ID: 23530).

Cluster 5, BP, Ef.
Overrepresented GO biological process terms among parasite genes which have low abundance
in sporulated oocysts, sporozoites and on day 7 p.i.. These genes seem to increase in abundance
upon infection and decrease late in infection, i.e., on day 7. In total the cluster contains 183 genes
of which 83 have GO annotations, and 100 do not.\\
 \\
\input{../cluster5_BP_Ef_OoSpoDown_day5Up.tex}

@@

Cluster 6, BP, Ef.
Overrepresented GO biological process terms among parasite genes which are highly abundant in sporozoites
but mainly downregulated in sporulated oocysts, apart from a few genes in the cluster.
This cluster seems to distinguish sporozoites from sporulated oocysts and have an average mRNA abundance in
all other samples. In total the cluster contains 180 genes of which 70 have GO annotations, and 110 do not.\\
 \\
\input{../cluster6_BP_Ef_SporozoitesUp.tex}

@@

Cluster 7, BP, Ef.
Overrepresented GO biological process terms among parasite genes which have low abundance in sporulated oocysts.
In sporozoites and early in infection, i.e., day 3 and 5 p.i., some of these genes are highly abundant. On day
7 p.i. these mRNAs have a below average abundance. In total the cluster contains 263 genes of which 131 have 
GO annotations, and 132 do not.\\
 \\
\input{../cluster7_BP_Ef_Day7OocyDown_SporoUP.tex}

@@

\section{Results for mouse}
  Cluster 1, BP, Mm
  Overrepresented GO biological process terms among mouse genes which have low abundance pre infection and 
  become more abundant in most day 3 and 5 samples from immunecompetent mice, but less in immune deficient Rag1-/- mice. 
  The upregulation trend is most clear in three first infection day 5 samples, which also form a separate
  sample cluster. On day 7, most of these genes are not differentially abundant, or below average, i.e., they
  seem to peak on day 5 p.i.. In total the cluster contains 273 genes of which 244 have GO annotations, and 29 
  do not.\\
   \\
  \input{../cluster1_BP_Mm_EarlyUp.tex}

@@

  Cluster 2, BP, Mm.
  Overrepresented GO biological process terms among mouse genes which are downregulated on day 7 p.i. and
  upregulated pre infection, i.e., these genes become more and more downregulated during infection.
  In total the cluster contains 198 genes of which 183 have GO annotations, and 15 do not. \\
  \input{../cluster2_BP_Mm_7dpiUp.tex}

@@

  Cluster 3, BP, Mm.
  Overrepresented GO biological process terms among mouse genes which are upregulated pre infection and
  downregulated on day 7 p.i., except in one NMRI second infection sample, possibly indicating no or
  poor infection in this sample. As in cluster 2, these mRNAs are most highly abundant before infection and
  become less abundant over the course of infection. In total the cluster contains 465 genes of which 440
  have GO annotations, and 25 do not\\
  \input{../cluster3_BP_Mm_7dpiDown.tex}

@@

  Cluster 4, BP, Mm.
  Overrepresented GO biological process terms among mouse genes which are highly abundant on day 7 p.i.. 
  There is overall no clear difference between pre infection and days 3 or 5 among these genesi. A slight
  tendency to higher abundance in second infection samples can be seen among these genes.
  In total the cluster contains 377 genes of which 360 have GO annotations, and 17 do not\\
  \input{../cluster4_BP_Mm_allInfDown.tex}
